Neural temporal point process for modeling laboratory measurements in ICU

### 2023-01-12 generic-color-brief.tex
first commit

\documentclass[journal,twoside,web]{ieeecolor}
\usepackage{generic}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}

\usepackage{multirow}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\markboth{\journalname, VOL. XX, NO. XX, XXXX 2017}
{Author \MakeLowercase{\textit{et al.}}: Point-process based representation learning for Electronic Health Records}
\begin{document}
\title{Point-process based representation learning for Electronic Health Records}
\author{Hojjat Karami, \IEEEmembership{Fellow, IEEE}, Anisoara Ionescu, and David Atienza, \IEEEmembership{Member, IEEE}
\thanks{This paragraph of the first footnote will contain the date on 
which you submitted your paper for review. It will also contain support 
information, including sponsor and financial support acknowledgment. For 
example, ``This work was supported in part by the U.S. Department of 
Commerce under Grant BS123456.'' }
\thanks{The next few paragraphs should contain 
the authors' current affiliations, including current address and e-mail. For 
example, F. A. Author is with the National Institute of Standards and 
Technology, Boulder, CO 80305 USA (e-mail: author@boulder.nist.gov). }
\thanks{S. B. Author, Jr., was with Rice University, Houston, TX 77005 USA. He is 
now with the Department of Physics, Colorado State University, Fort Collins, 
CO 80523 USA (e-mail: author@lamar.colostate.edu).}
\thanks{T. C. Author is with 
the Electrical Engineering Department, University of Colorado, Boulder, CO 
80309 USA, on leave from the National Research Institute for Metals, 
Tsukuba, Japan (e-mail: author@nrim.go.jp).}}

@@

\begin{IEEEkeywords}
Enter keywords or phrases in alphabetical 
order, separated by commas. For a list of suggested keywords, send a blank 
e-mail to keywords@ieee.org or visit the site.
\end{IEEEkeywords}

\section{Introduction}
\label{sec:intro}

Writing an effective abstract is an indispensable part of any form of research, since it that can motivate the audience to read follow the rest of the text. In this task, I have analyzed five abstracts from the field of artificial intelligence (AI). The table in section 2 shows a summary of the most important aspects of the analyzed abstracts.

@@

\textbf{Temporal point process}

\textbf{Neural temporal point process}

\textbf{Handling irregular sampling}

@@

\begin{align*}
    f^{\prime}(\mathcal{S}_p) = g^{\prime} \left(  \frac{1}{p} \sum_{s_k \in \mathcal{S}_p}  h^{\prime}_{\theta}(s_k) \right)  \\
    K_p=[f^{\prime}(\mathcal{S}_p), s_p]^T W^K   
\end{align*}

Furthermore, we define a query vector $w^q \in \mathbb{R}^d$, which allows the model to summarize different aspects of the dataset via

@@

Without loss of generality, we can consider multiple heads by adding an additional dimension to keys and queries.\\

@@

\subsection{Mark and Time Decoder}

by concatenating encoded events $x_{enc}$ and additional information $f^{\prime}(\mathcal{S}_p)$, we can predict next marks and times as follows:

\begin{equation}
    \hat{e}_{j+1}=MLP([x_j, f^{\prime}(\mathcal{S}_j)])
\end{equation}

\begin{equation}
    \hat{t}_{j+1}=MLP([x_j, f^{\prime}(\mathcal{S}_j)])
\end{equation}

\begin{equation}
    \mathcal{L}_{mark} =
\end{equation}
\begin{equation}
    \mathcal{L}_{time} = \sum_{j=1}^{L-1} \left(  (\hat{t}_{j+1}-t_j)-(t_{j+1}-t_j)  \right)^2
\end{equation}

\subsection{Event Decoder}
Once we obtain a representation of a patient using embedded events and states, we can try to parameterize conditional intensity functions (CIFs) of the events.

In neural point process literature, many approaches have been propose to decode either conditional or cumulative intensity function.

\begin{equation}
    \lambda_k(t|\mathcal{H}_t ) = f_k\left(\alpha_k \frac{t-t_j}{t_j}+\mathbf{w}_k^T \mathbf{x}_{enc}(t_j)+\mathbf{y}_k^T \mathbf{s}_{enc}(t_j) +b_k   \right)
\end{equation}

\begin{equation}
    \lambda_k(t|\mathcal{H}_t ) = f_k\left(\alpha_k \frac{t-t_j}{t_j}+\mathbf{w}_k^T \mathbf{x}_{enc}(t_j)+b_k   \right)
\end{equation}

@@

\textbf{Physionet 2019 Sepsis Early Prediction Challenge (\emph{P19})}.
This dataset contains clinical data of about 40k patients in ICU. Clinical data consist of demographics, vital signs and laboratory values as well as sepsis label in a one-hour time grid. Our objective is to predict the timestamp of next lab sampling events as well as measured variables (event marks) given the patient history.

% To show the effectiveness of the proposed method, we further consider the following scenarios:
% \begin{itemize}
%     \item Enc(Event)+Dec(NextEvent): Here
%     \item Enc(Event)+Dec(NextEvent+CIF): Here
%     \item Enc(Event+state)+Dec(NextEvent): Here
%     \item Enc(Event+state)+Dec(NextEvent+CIF): Here
% \end{itemize}

\textbf{MIMIC-IV (\emph{M4})}.
We selected Medical Information Mart for Intensive Care (MIMIC) IV [1], which is a real-world clinical database comprising health data relating to over 40,000 patients admitted to ICU at the Beth Israel Deaconess Medical Center.

\textbf{Synthea (\emph{Syn})}.
We used the Synthea simulator (Walonoski et al., 2018) which generates patient-level EHRs using human expert curated Markov processes. Here, we reused the already processed version of this data by Edgauard.

\textbf{Stackoverflow (\emph{SO})}.
StackOverflow is a question-answering website. The website rewards users with badges to promote engagement in the community, and the same badge can be rewarded multiple times to the same user. We collect data in a two-year period, and we treat each user’s reward history as a sequence. Each event in the sequence signifies receipt of a particular medal.

\subsection*{Scenarios}
To investigate the effectiveness of the proposed method, we consider three input scenarios (TE, DA, TE+DA) as well as three loss functions (next event, CIF, next event+CIF) which would result in nine scenarios.

The first series of experiments are conducted to investigate the advantage of encoding additional information for paramterization of intensity functions. We consider seven scenarios:
Here, the baseline models

To show the effectiveness of time concatenation we report.

\subsection*{Baselines}
we use NEURALTPP that is already developed pipeline by as they already considered a lot of comibinations.

\subsection*{Metrics}
We report the weighted AUPRC, AUROC of next predicted event as well as root mean sqaure error (RMSE) of next measurement interval.
For evaluating the goodness of fit for the parameterized point process, we report normalized negative likelihood normalized by number of ocurred event (NLL/events). Furthermore, we can also evaluate the learned representation of each patient to predict the sepsis label in a binary classification task.

\section{Results and Discussion}
\label{sec:Results and Discussion}

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[h!]
    \centering
    \caption{unsupervised learning}
    \label{tab:table1}
    \resizebox{\columnwidth}{!}{%    
    \begin{tabular}{cclllll}
    \hline
    \multirow{2}{*}{Dataset} & \multicolumn{2}{c}{Model}            & \multicolumn{2}{c}{next mark} & next time & CIFs  \\
                             & \multicolumn{1}{l}{module} & loss    & AUPRC         & AUROC         & RMSE      & NLL   \\ \hline
    \multirow{6}{*}{p12}     & \multirow{3}{*}{TE}        & MC      & 23.98         & 23.98         & 23.98     & 23.98 \\
                             &                            & ML      & 23.98         & 23.98         & 23.98     & 23.98 \\
                             &                            & SHPMARK & 23.98         & 23.98         & 23.98     & 23.98 \\
                             & \multirow{3}{*}{TEDA}      & MC      & 23.98         & 23.98         & 23.98     & 23.98 \\
                             &                            & ML      & 23.98         & 23.98         & 23.98     & 23.98 \\
                             &                            & SHPMARK & 23.98         & 23.98         & 23.98     & 23.98 \\ \hline
    \multirow{6}{*}{p19}     & \multirow{3}{*}{TE}        & MC      & 23.98         & 23.98         & 23.98     & 23.98 \\
                             &                            & ML      & 23.98         & 23.98         & 23.98     & 23.98 \\
                             &                            & SHPMARK & 23.98         & 23.98         & 23.98     & 23.98 \\
                             & \multirow{3}{*}{TEDA}      & MC      & 23.98         & 23.98         & 23.98     & 23.98 \\
                             &                            & ML      & 23.98         & 23.98         & 23.98     & 23.98 \\
                             &                            & SHPMARK & 23.98         & 23.98         & 23.98     & 23.98 \\ \hline
    \end{tabular}%
    }
\end{table}

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[h!]

    \centering
    \caption{Transfer learnings}
    \label{tab:table2}
    \resizebox{\columnwidth}{!}{%
    \begin{tabular}{|cll|ll|ll|}
    \hline
    \multicolumn{3}{|c|}{model}                                                          & \multicolumn{2}{c|}{p12}           & \multicolumn{2}{c|}{p19}           \\ \hline
    \multicolumn{1}{|l|}{module}                & \multicolumn{1}{l|}{loss}    & freezed & \multicolumn{1}{l|}{AUPRC} & AUROC & \multicolumn{1}{l|}{AUPRC} & AUROC \\ \hline
    \multicolumn{1}{|c|}{\multirow{3}{*}{TE}}   & \multicolumn{1}{l|}{MC}      &         & \multicolumn{1}{l|}{23.98} & 23.98 & \multicolumn{1}{l|}{23.98} & 23.98 \\ \cline{2-7} 
    \multicolumn{1}{|c|}{}                      & \multicolumn{1}{l|}{ML}      &         & \multicolumn{1}{l|}{23.98} & 23.98 & \multicolumn{1}{l|}{23.98} & 23.98 \\ \cline{2-7} 
    \multicolumn{1}{|c|}{}                      & \multicolumn{1}{l|}{SHPMARK} &         & \multicolumn{1}{l|}{23.98} & 23.98 & \multicolumn{1}{l|}{23.98} & 23.98 \\ \hline
    \multicolumn{1}{|c|}{\multirow{3}{*}{TEDA}} & \multicolumn{1}{l|}{MC}      &         & \multicolumn{1}{l|}{23.98} & 23.98 & \multicolumn{1}{l|}{23.98} & 23.98 \\ \cline{2-7} 
    \multicolumn{1}{|c|}{}                      & \multicolumn{1}{l|}{ML}      &         & \multicolumn{1}{l|}{23.98} & 23.98 & \multicolumn{1}{l|}{23.98} & 23.98 \\ \cline{2-7} 
    \multicolumn{1}{|c|}{}                      & \multicolumn{1}{l|}{SHPMARK} &         & \multicolumn{1}{l|}{23.98} & 23.98 & \multicolumn{1}{l|}{23.98} & 23.98 \\ \hline
    \end{tabular}%
    }
    \end{table}

\subsection{Next event prediction}

using CIF does not seem to improve the next mark and time prediction.
this is new insight as previous studies has claimed this loss to be useful.

\subsection{Learned representation}

using CIF loss function leads to a better representation for sepsis.

\subsection{Likelihood estimation}

Although CIF does not imporve mark prediction, it has led to better representation of patient for downstream task such as sepsis prediction.

In addition, we can interprete some of learned CIF patterns.

explain the effect of time concatenation in SO dataset

\subsection{Model interpretability}

one advantage of proposed method is use of attention mechanisms in both event and state encoder.

tsne of learned representation. 4 modes:

\begin{itemize}
    \item (DA,TE)->(Mark, CIF)
\end{itemize}

attention of DA for sepsis prediction

attention matrix of events for SO dataset

@@

\section{Introduction}
\label{sec:introduction}
\IEEEPARstart{T}{his} document is a template for \LaTeX. If you are 
reading a paper or PDF version of this document, please download the 
electronic file, trans\_jour.tex, from the IEEE Web site at \underline
{http://www.ieee.org/authortools/trans\_jour.tex} so you can use it to prepare your manuscript. If 
you would prefer to use LaTeX, download IEEE's LaTeX style and sample files 
from the same Web page. You can also explore using the Overleaf editor at 
\underline
{https://www.overleaf.com/blog/278-how-to-use-overleaf-with-}\discretionary{}{}{}\underline
{ieee-collabratec-your-quick-guide-to-getting-started\#.}\discretionary{}{}{}\underline{xsVp6tpPkrKM9}

@@

\subsection{Multipart figures}
Figures compiled of more than one sub-figure presented side-by-side, or 
stacked. If a multipart figure is made up of multiple figure
types (one part is lineart, and another is grayscale or color) the figure 
should meet the stricter guidelines.

@@

\subsection{Review Stage Using ScholarOne\textregistered\ Manuscripts}
Contributions to the Transactions, Journals, and Letters may be submitted 
electronically on IEEE's on-line manuscript submission and peer-review 
system, ScholarOne\textregistered\ Manuscripts. You can get a listing of the 
publications that participate in ScholarOne at 
\underline{http://www.ieee.org/publications\_standards/publications/}\discretionary{}{}{}\underline{authors/authors\_submission.html}.
First check if you have an existing account. If there is none, please create 
a new account. After logging in, go to your Author Center and click ``Submit 
First Draft of a New Manuscript.''

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{a1.png}}]{First A. Author} (M'76--SM'81--F'87) and all authors may include 
biographies. Biographies are often not included in conference-related
papers. This author became a Member (M) of IEEE in 1976, a Senior
Member (SM) in 1981, and a Fellow (F) in 1987. The first paragraph may
contain a place and/or date of birth (list place, then date). Next,
the author's educational background is listed. The degrees should be
listed with type of degree in what field, which institution, city,
state, and country, and year the degree was earned. The author's major
field of study should be lower-cased.

The second paragraph uses the pronoun of the person (he or she) and not the 
author's last name. It lists military and work experience, including summer 
and fellowship jobs. Job titles are capitalized. The current job must have a 
location; previous positions may be listed 
without one. Information concerning previous publications may be included. 
Try not to list more than three books or published articles. The format for 
listing publishers of a book within the biography is: title of book 
(publisher name, year) similar to a reference. Current and previous research 
interests end the paragraph. The third paragraph begins with the author's 
title and last name (e.g., Dr.\ Smith, Prof.\ Jones, Mr.\ Kajor, Ms.\ Hunter). 
List any memberships in professional societies other than the IEEE. Finally, 
list any awards and work for IEEE committees and publications. If a 
photograph is provided, it should be of good quality, and 
professional-looking. Following are two examples of an author's biography.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{a2.png}}]{Second B. Author} was born in Greenwich Village, New York, NY, USA in 
1977. He received the B.S. and M.S. degrees in aerospace engineering from 
the University of Virginia, Charlottesville, in 2001 and the Ph.D. degree in 
mechanical engineering from Drexel University, Philadelphia, PA, in 2008.

From 2001 to 2004, he was a Research Assistant with the Princeton Plasma 
Physics Laboratory. Since 2009, he has been an Assistant Professor with the 
Mechanical Engineering Department, Texas A{\&}M University, College Station. 
He is the author of three books, more than 150 articles, and more than 70 
inventions. His research interests include high-pressure and high-density 
nonthermal plasma discharge processes and applications, microscale plasma 
discharges, discharges in liquids, spectroscopic diagnostics, plasma 
propulsion, and innovation plasma applications. He is an Associate Editor of 
the journal \emph{Earth, Moon, Planets}, and holds two patents.

Dr. Author was a recipient of the International Association of Geomagnetism 
and Aeronomy Young Scientist Award for Excellence in 2008, and the IEEE 
Electromagnetic Compatibility Society Best Symposium Paper Award in 2011. 
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{a3.png}}]{Third C. Author, Jr.} (M'87) received the B.S. degree in mechanical 
engineering from National Chung Cheng University, Chiayi, Taiwan, in 2004 
and the M.S. degree in mechanical engineering from National Tsing Hua 
University, Hsinchu, Taiwan, in 2006. He is currently pursuing the Ph.D. 
degree in mechanical engineering at Texas A{\&}M University, College 
Station, TX, USA.

From 2008 to 2009, he was a Research Assistant with the Institute of 
Physics, Academia Sinica, Tapei, Taiwan. His research interest includes the 
development of surface processing and biological/medical treatment 
techniques using nonthermal atmospheric pressure plasmas, fundamental study 
of plasma sources, and fabrication of micro- or nanostructured surfaces.

Mr. Author's awards and honors include the Frew Fellowship (Australian 
Academy of Science), the I. I. Rabi Prize (APS), the European Frequency and 
Time Forum Award, the Carl Zeiss Research Award, the William F. Meggers 
Award and the Adolph Lomb Medal (OSA).
\end{IEEEbiography}

>>>>

\documentclass[9pt,shortpaper,twoside,web]{ieeecolor}
\usepackage{generic}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}

\usepackage{amsmath}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\markboth{\journalname, VOL. XX, NO. XX, XXXX 2017}
{Author \MakeLowercase{\textit{et al.}}: Neural temporal point process for
modeling laboratory measurements in ICU}
\begin{document}
\title{Neural temporal point process for modeling laboratory measurements in ICU}
\author{H. Karami, \IEEEmembership{Fellow, IEEE}, Second B. Author, and Third C.
Author, Jr., \IEEEmembership{Member, IEEE} 
\thanks{This paragraph of the first
footnote will contain the date on which you submitted your paper for review. It
will also contain support information, including sponsor and financial support
acknowledgment. For example, ``This work was supported in part by the U.S.
Department of Commerce under Grant BS123456.'' } \thanks{The next few paragraphs
should contain the authors' current affiliations, including current address and
e-mail. For example, F. A. Author is with the National Institute of Standards
and Technology, Boulder, CO 80305 USA (e-mail: author@boulder.nist.gov). }
\thanks{S. B. Author, Jr., was with Rice University, Houston, TX 77005 USA. He
is now with the Department of Physics, Colorado State University, Fort Collins,
CO 80523 USA (e-mail: author@lamar.colostate.edu).} \thanks{T. C. Author is with
the Electrical Engineering Department, University of Colorado, Boulder, CO 80309
USA, on leave from the National Research Institute for Metals, Tsukuba, Japan
(e-mail: author@nrim.go.jp).}}

@@

\begin{IEEEkeywords}
point process, Electronic Health Records (EHR),
\end{IEEEkeywords}

\section{Introduction}
\label{sec:introduction}

@@

@@

\begin{align*}
    f^{\prime}(a\mathcal{S}_p) = g^{\prime} \left(  \frac{1}{p} \sum_{s_k \in \mathcal{S}_p}  h^{\prime}_{\theta}(s_k) \right)  \\
    K_p=[f^{\prime}(\mathcal{S}_p), s_p]^T W^K   
\end{align*}

Furthermore, we define a query vector $w^q \in \mathbb{R}^d$, which allow the model to summarize different aspects of the dataset via

@@

@@

@@

\textbf{Physionet 2019 Sepsis Early Prediction Challenge }.
This is a dataset contains clinical data of about 40k patients in ICU. Clinical data consist of demographics, vital signs and laboratory values as well as sepsis label in a one-hour time grid. Our objective is to predict the timestamp of next lab sampling events as well as measured variables (event marks) given the patient history. fsdfs.

\subsection*{Baselines}

\section{Results and Discussion}

\label{sec:Results and Discussion}

@@

\begin{equation}

\frac{1}{n} \newline
\frac{d^2 y}{d x^2} \bigotimes \cos \mathcal{L} \mathbb{L} \hat{\rightarrow \Rightarrow } 
\Rightarrow \Longrightarrow 
\end{equation}

This document is a template for \LaTeX. If you are 
reading a paper or PDF version of this document, please download the 
electronic file, trans\_jour.tex, from the IEEE Web site at \underline
{http://www.ieee.org/authortools/trans\_jour.tex} so you can use it to prepare your manuscript. If 
you would prefer to use LaTeX, download IEEE's LaTeX style and sample files 
from the same Web page. You can also explore using the Overleaf editor at 
\underline
{https://www.overleaf.com/blog/278-how-to-use-overleaf-with-}\discretionary{}{}{}\underline
{ieee-collabratec-your-quick-guide-to-getting-started\#.}\discretionary{}{}{}\underline{xsVp6tpPkrKM9}

@@

\subsection{Multipart figures}
Figures compiled of more than one sub-figure are presented side-by-side, or 
stacked. If a multipart figure is made up of multiple figure
types (one part is lineart, and another is grayscale or color) the figure 
should meet the stricter guidelines.

@@

\subsection{Review Stage Using ScholarOne\textregistered\ Manuscripts}
Contributions to the Transactions, Journals, and Letters may be submitted 
electronically on IEEE's online manuscript submission and peer-review 
system, ScholarOne\textregistered\ Manuscripts. You can get a listing of the 
publications that participate in ScholarOne at 
\underline{http://www.ieee.org/publications\_standards/publications/}\discretionary{}{}{}\underline{authors/authors\_submission.html}.
First check if you have an existing account. If there is none, please create 
a new account. After logging in, go to your Author Center and click ``Submit 
First Draft of a New Manuscript.''

@@
